---
title: "HW4"
author: "Malvika Mitra" 
date: "8th Oct 2018" 
output:
  github_document:
    toc: yes 
---

```{r load libraries}
suppressPackageStartupMessages(library(tidyverse)) # to supress messages after library calls
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(gapminder))
library(knitr)
library(dplyr)
```

# DATA RESHAPING:

Data Reshaping is about changing the way data is organized into rows and columns. Most of the time data processing is done by taking the input data as a data frame. It is easy to extract data from the rows and columns of a data frame but there are situations when we need the data frame in a format that is different from format in which we received it. Functions like gather() and spread() from **tidyr** reshape the rows to columns in a data frame.

## Activity #2

Make a tibble with one row per year and columns for life expectancy for two or more countries.

Use knitr::kable() to make this table look pretty in your rendered homework.

Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.

```{r}
my_gap<-gapminder %>% #making a new dataframe and saving it                         in the variable my_gap.
  group_by(year) %>% 
  filter((country=="Canada"| country=="Algeria")) 
my_gap

g<-tibble(year=my_gap$year,lf=my_gap$lifeExp,country=my_gap$country)  #making a tibble
knitr::kable(g)

gapminder %>% 
  filter((country=="Canada"| country=="Algeria")) %>% 
ggplot(aes(year,lifeExp))+
  geom_point(aes(color=country))+
  ggtitle("      ScatterPlot of LifeExp")+
  xlab("Year")+
  ylab("LifeExp")

```
Output: Filtered the Gapminder data according to requirements and saved it my_gap variable.

Then made a tibble out of the columns of my_gap.

Knitted the table and finally made the scatter plot of the lifeExp of Canada ana Algeria over the years.

The graph shows that the lifeExp of Canada is higher than that of Algeria.

## Activity #3

Compute some measure of life expectancy (mean? median? min? max?) for all possible combinations of continent and year. Reshape that to have one row per year and one variable for each continent. Or the other way around: one row per continent and one variable per year.

Use knitr::kable() to make these tables look pretty in your rendered homework.

Is there a plot that is easier to make with the data in this shape versis the usual form? If so (or you think so), try it! Reflect.

```{r}
k<-gapminder %>% 
   group_by(year,continent) %>% 
  summarise(medianLE=median(lifeExp))

knitr::kable(k)

ggplot(k,aes(continent,medianLE))+
  geom_point()

n<-spread(k,key="continent",value="year")
n
ggplot(n,aes(Africa,medianLE))+
  geom_point()

ggplot(n,aes(Asia,medianLE))+
  geom_point()
```
Output:

Grouped by continent and year.

Used the summarise() fn to calculate the median lifeExp.

Used spread() fn to spread the data continent wise according to the year.

I dont think its easier because we will have to make separate plots for each continent versus earlier we had one plot with all the continents and their respective medianLF plotted on it.

## Activity #4

In Window functions, we formed a tibble with 24 rows: 2 per year, giving the country with both the lowest and highest life expectancy (in Asia). Take that table (or a similar one for all continents) and reshape it so you have one row per year or per year * continent combination.

```{r}
a<-my_gap %>%
  filter(continent == "Asia") %>%
  select(year, country, lifeExp) %>%
  group_by(year) %>%
  filter(min_rank(desc(lifeExp)) < 2 | min_rank(lifeExp) < 2) %>% 
  arrange(year) %>%
  print(n = Inf)


my_gap
spread(a,country,year)
  
```
Output:

Reshaped the data.

## Activity #5 

Previous TA Andrew MacDonald has a nice data manipulation sampler. Make up a similar set of exercises for yourself, in the abstract or (even better) using Gapminder or other data, and solve them.

Creating my dataframe
```{r}
A <- data.frame(
       x1 = c('A', 'A','A', 'B', 'B'),
       x2 = c('a', 'b', 'b', 'c', 'd'),
       x3 = c(1, 1, 1, 2, 2))
A
```

GROUP

```{r}
B<-A %>% 
  group_by(x1,x2) %>% 
  summarize(Nrows=n(),
            mean=mean(x3))
B
```
Output:

Created a dataframe B and grouped by columns x1 and x2.
the Nrows coulmns shows the count of the rows under the corresponding values of x1 and x2.Example A of x1 has a corresponding value of a in just one row of x2 so Nrows value is 1 for this entrie and mean column just calulates the mean of x3 values corresponding to x1 values.

SUBSET

subset Table A to give C:

```{r}
C<-filter(A,x1=="A")
C
```


TRANSPOSE

transpose C to give:
```{r}
C_transpose <- data.frame(t(C))
C_transpose
```

SORT

```{r}
A_arranged <- arrange(A,x1,x2,x3)
A_arranged
```



# Join Prompts

Join Prompts are used to combine datasets in R, using various functions like left_join,right_join,inner_join,full_join etc. 

## Activity #1

Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a dplyr join function and make some observations about the process and result. Explore the different types of joins. Examples of a second data frame you could build:
 One row per country, a country variable and one or more variables with extra info, such as language spoken, NATO membership, national animal, or capitol city.
 One row per continent, a continent variable and one or more variables with extra info, such as northern versus southern hemisphere.

```{r}
gap<-gapminder[1:5,] %>% 
  select(country,year,lifeExp)
gap
cont<-data.frame(
  year=c(1952,1957,1962,1967,1972),
  gdp=c(779.4453,820.8530,853.1007,836.1971,739.9811)
)
cont

left_join(gap,cont)

inner_join(gap,cont)

semi_join(gap,cont)

full_join(gap,cont)
```
Output:

left_join: Returns all rows of gap, and all columns of gap and cont.

inner_join:Return all rows of gap where there are matching values in cont, and all columns of gap and cont.

semi_join: Returns all rows of gap where there are matching values in cont, keeping just columns from gap. A semi join differs from an inner join because an inner join will return one row of gap for each matching row of cont, where as a semi join will never duplicate rows of gap. 


full_join: Returns all rows and all columns from both gap and cont. 

## Activity #2

Create your own cheatsheet patterned after Jenny’s but focused on something you care about more than comics! Inspirational examples:

Pets I have owned + breed + friendly vs. unfriendly + ??. Join to a table of pet breed, including variables for furry vs not furry, mammal true or false, etc.
Movies and studios….
Athletes and teams….

You will likely need to iterate between your data prep and your joining to make your explorations comprehensive and interesting. For example, you will want a specific amount (or lack) of overlap between the two data.frames, in order to demonstrate all the different joins. You will want both the data frames to be as small as possible, while still retaining the expository value.

```{r}
pet<- data.frame(
  name=c("tim","lily","jin"),
  type=c("dog","parrot","cat"),
  nature=c("friendly","friendly","nonfriendly")
)
  animal<-data.frame(
    type=c("dog","bird","cat"),
    life=c(10,2,9)
    )
  
pet
animal
```
Ctreated two dataframes pet and animal and will now explore various joins on them.

left_join(pet,animal):

```{r}
left_join(pet,animal,by="type")
```
Output:

Join matching rows from `animal` to `pet` by matching "type" variable.

The type column of pet is printed and since parrot does mot have a matching value in animal dataframe a NA is printed in the life column.

right_join(pet,animal):

```{r}
right_join(pet,animal,by="type")
```
Output:

Join matching rows from `animal` to `pet` by matching "type" variable.

The type column of animal is printed and since bird does mot have a matching value in animal dataframe a NA is printed in both name and nature columns.

inner_join(pet,animal):

```{r}
inner_join(pet,animal)
```
Output:

Returns all rows from pet where there are matching values in animal, and all columns from pet and animal.

We lose parrot in the join because, although he appears in  pet, but does not appear in animal. The join result has all variables from pet plus life, from animal.

semi_join(pet,animal):

```{r}
semi_join(pet,animal) 
```
Output:

Returns all rows from pet where there are matching values in animal, keeping just columns from pet. 
A semi join differs from an inner join because an inner join will return one row of pet for each matching row of animal, whereas a semi join will never duplicate rows of pet. This is a filtering join.

anti_join(pet,animal):

```{r}
anti_join(pet,animal)
```
Output:

Return all rows from pet where there are not matching values in animal, keeping just columns from pet. This is a filtering join.

left_join(animal,pet):

```{r}
left_join(animal,pet)
```
Output:

Join matching rows from `pet` to `animal` by matching "type" variable.

The type column of animal is printed and since bird does mot have a matching value in pet dataframe a NA is printed in the name and nature columns.

right_join(animal,pet):

```{r}
right_join(animal,pet)
```
Output:

Join matching rows from `pet` to `animal` by matching "type" variable.

The type column of pet is printed and since parrot does mot have a matching value in animal dataframe a NA is printed in the life column.


inner_join(animal,pet):

```{r}
inner_join(animal,pet)
```

Output:
Returns all rows from animal where there are matching values in pet, and all columns from pet and animal.

We lose bird in the join because, although he appears in animal, but does not appear in pet. The join result has life from animal plus name, nature from pet.

semi_join(animal,pet):

```{r}
semi_join(animal,pet)
```
Output:

Returns all rows from animal where there are matching values in pet, keeping just columns from animal. 
A semi join differs from an inner join because an inner join will return one row of animal for each matching row of pet, whereas a semi join will never duplicate rows of animal. This is a filtering join.

anti_join(animal,pet):

```{r}
anti_join(animal,pet)
```
Output:

Return all rows from animal that have no matching values in pet, keeping just columns from animal. This is a filtering join.

full_join(pet,animal):

```{r}
full_join(pet,animal)
```
Output:
 Return all rows and all columns from both pet and animal. Where there are not matching values, returns NA for the one missing. This is a mutating join.
 
We get all variables from pet AND all variables from animal. Any row that derives solely from one table or the other carries NAs in the variables found only in the other table.
 
```{r}
animal2<-data.frame(
    type=c("tiger","bird","lion"),
    life=c(10,5,9)
    )
```

left_joint(pet,animal2):

```{r}
left_join(pet,animal2)
```
Output:

Joins matching rows from animal2 to pet.

Since the type column values dont match and the life times of a dog, parrot and cat are not defined in the type column of animal2 thats why after joining the two datasets the joined life column gives NAs.

inner_join(pet,animal2):

```{r}
inner_join(pet,animal2)
```
Output:

Joins data. Retains only rows common in both sets.
Since there are no common rows so it returns 0 rows in the result.

anti_join():

```{r}
anti_join(pet,animal2)
```
Output:

Returns all rows in pet that do not have a match in animal.


full_join(pet,animal2):

```{r}
full_join(pet,animal2)
```
Output:

Joins the data from both the dataframes.Retains all values,all rows from both.
